Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2023-12-19 2024pipit/section/20-Motivation.tex
ok

Talk about the problem.

% \newcommand\signal{\textit{signal}}
\begin{tabbing}
  @MM@\= @MM@ \= \kill
  @let@ count_when ($\textit{limit}$: $\NN$) ($\textit{inc}$: stream $\BB$): stream $\NN$ = \\
    \> @let rec@ $\textit{count}$ @=@ \\
    \> \> @if@ $\textit{count} >= \textit{limit}$ @then@ $\textit{limit}$ \\
    \> \> @else@ $(0~@fby@~\textit{count}) + (@if@~\textit{inc}~@then@~1~@else@~0)$ \\
    \> @in@ $\textit{count}$ \\
\end{tabbing}

\begin{tabbing}
  @MM@\= @MM@ \= \kill
  @let@ trigger_fetch ($\textit{index}$: stream $\NN$): stream trigger = \\
    \> @contract@ \\
    \> \> \{ $0 \le \textit{index} \le \textit{maximum}$ \} \\
    \> \> Array.index_const \textit{triggers_array} \textit{index} \\
    \> \> \{ trigger.~$\true$ \} \\
\end{tabbing}

>>>>

Pure streaming languages like Lustre lie in an uneasy space somewhere between pure functions and imperative programs.
Like pure functions, pure streaming languages are equational and support reasoning by unfolding and rewriting.
Like imperative programs, they describe how state evolves over time.
Reasoning about them is somewhat tricky, then, as one must reason about finite prefixes of streams.

As a concrete example to introduce Pipit, we consider a program with a static schedule of \emph{triggers}, or actions to be performed at a particular time.
These triggers are stored in a constant array.
The program maintains an index that refers to the current active trigger; at each instant in time, the program checks if the active trigger has expired, and if so, it increments the index.
This trigger logic is a simplification of the time-triggered CAN bus specification \cite{fuehrer2001time}; we will discuss further in \autoref{s:evaluation}.

We first implement a streaming function \emph{count_when} to maintain the index; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
At each time step, $\mbox{count_when}~\textit{max}~\textit{inc}$ checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the previous counter as-is.

% \newcommand\signal{\textit{signal}}
\begin{tabbing}
  @MM@\= @MM@ \= \kill
  @let@ count_when ($\textit{max}$: $\NN$) ($\textit{inc}$: stream $\BB$): stream $\NN$ = \\
    \> $@rec@~\rawbind{\textit{count}}{}$ \\
    \> \> @let@ $\textit{count'}$ @=@ $(0~@fby@~\textit{count}) + (@if@~\textit{inc}~@then@~1~@else@~0)$ @in@ \\
    \> \> $\xcheckP{\PSUnknown}{(0 \le \textit{count} \le \textit{max})}$; \\
    \> \> @if @ $\textit{count'} \ge \textit{max}$ @then@ $\textit{max}$  @else@ $\textit{count'}$ \\
    % \> $\textit{count}$ \\
\end{tabbing}

The implementation of \emph{count_when} first defines a recursive stream, \emph{count}, which in turn defines a local incremented stream \emph{count'}.
The syntax $0~@fby@~\textit{count}$ can be read as zero \emph{followed by} the previous count.

The syntax $\xcheckP{\PSUnknown}{(0 \le \textit{count} \le \textit{max})}$ asserts that the count is within the range $[0, \textit{max}]$.
The subscript $\PSUnknown$ on the check is the \emph{property status}, which in this case denotes that the assertion has been stated, but it is not yet known whether it holds.
A property status of $\PSValid$, on the other hand, denotes that a property has been proved to hold.
These property statuses are used to defer checking properties until enough is known about the environment, and to avoid rechecking properties that have already been proven.

Pipit is an embedded domain-specific language. % we usually present Pipit programs using \fstar{} syntactic sugar rather than showing the desugared core program.
The program above is really syntactic sugar for a \fstar{} program that takes a natural number and constructs a Pipit core expression with a free boolean variable; we will discuss the details of the core language in \autoref{s:core}.

To actually prove the property above, we translate the program into a transition system and prove the property inductively on the system.
Finally, we \emph{bless} the expression, which marks the properties as valid.
Blessing is an intensional operation: it modifies the internal structure of the expression, but does not affect the operational semantics.

\begin{tabbing}
  @MM@\= @MM@ \= \kill
  @let@ $\mbox{count_when}_{\PSValid}$ ($\textit{max}$: $\NN$): stream $\BB$ $\to$ stream $\NN$ = \\
    \> @let@ $\textit{system} = \mbox{System.translate}_1 (\mbox{count_when}~\textit{max})$ @in@ \\
    \> @assert@ (System.inductive_check $\textit{system}$) @by@ (pipit\_simplify ()); \\
    \> $@bless@_1$ $(\mbox{count_when}~\textit{max})$
\end{tabbing}

The subscript 1 in the translation to transition system and blessing operations refers to the fact that the stream function has one parameter.
This proof does require some boilerplate; in the future we plan to minimise this with metaprogramming.

Callers of count_when can now use the valid variant without needing to re-prove the count-range property.

\begin{tabbing}
  @MM@\= @MM@ \= \kill
  @let@ trigger_fetch ($\textit{index}$: stream $\NN$): stream trigger = \\
    \> $@contract@_{\PSUnknown}$ \\
    \> \> \{ $0 \le \textit{index} \le \textit{maximum}$ \} \\
    \> \> Array.index_const \textit{triggers_array} \textit{index} \\
    \> \> \{ trigger.~$\true$ \} \\
\end{tabbing}
